---
title: "termiteN"
author: "Tristan Caro"
date: "11/24/2020"
output: html_document
code_folding: hide
header-includes:
  -\usepackage{textcomp}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(tidyverse)
library(reticulate)
```


# Termites

Many non-biological and biological processes create regularly patterned landscapes. However, there is not much understanding of how patterning impacts ecological behavior.

Spatial structure in biotic processes is an important factor in governing community-level interactions and ecosystem processes. In many cases, a keystone species exerts community-wide effects. Often, this effect is created through indirect interactions that are mediated through an abiotic system or process.

Termites have been identified as a keystone species in the African savannas. Termites are particularly interesting due to their ability to farm fungi.

Legumes, like **Acacia**, access atmospheric $N_2$ by cultivating symbiotic relationships with nitrogen-fixing bacteria. This can be an important input of N into N-limited ecosystems. Although levels of N fixation vary amongst species and across topoedaphic and cliatic gradients, the **Acacia** tree's ability to collect large quantities of N from fixation allows them to thrive in N-depleted soils.

Nitrogen isotope $\delta ^{15}N$ values measured in plant tissues can provide insight into the relative contributions of N sources (fixed N vs. soil N).

Termite mounds in the African savanna are nutrient-rich for a variety of reasons. 

# The 15N Natural Abundance Method

First, we can define isotopic values in delta notation as such:
$$
\delta^{15}N = \frac{R_{sample} - R_{std}}{R_{std}} * 1000 ‰
$$
Where R is the ratio of isotope $^{15}N$ "heavy" to $^{14}N$ "light":
$$
R = (^{15}N/^{14}N)
$$

The percentage of 15N from the atmosphere out of the total N in a plant sample ($\% N_{dfa}$) is calculated as below:

$$
\%N_{dfa} = (\delta^{15}N_0 - \delta^{15}N_t) / (\delta^{15}N_0 - \delta^{15}N_a) \space * 100 \%
$$
where 
$\delta^{15}N_0$ : a non-N2 fixing reference plant, in our case, **Aerva**
$\delta^{15}N_t$ : a N2 fixing plant, in our case **Acacia**
$\delta^{15}N_a$ : N2 fixing plants in an N-free medium


We can think about the Nitrogen isotopic composition of the **Acacia** tree using a simple mixing model that is a function of proportion $N_2 fixation$:

$$
\delta^{15}N_{acacia} = pN_{fix}(\delta^{15}N_{atm} + \Delta^{15}_{assim - atm}) \\
+ (1 - pN_{fix})(\delta^{15}N_{soil} + \Delta^{15}N_{assim-soil})
$$

Where:

- $pN_{fix}$ : the proportion of **Acacia** leaf N that is derived from fixation
- $\Delta^{15}N_{assim}$ : isotopic spacing from the source (atmospheric N to plant-available soil N) to leaf tissue.
- $\Delta^{15}N_{assim}$ : an approximation of the actual isotopic differences between the sources and leaf tissue.
Values for $\delta^{15}N_{atm}$ and $\delta^{15}N_{soil}$ are the values for atmospheric $N_2$ and soil-derived plant-available N, respectively. 

In essence, using mass-balance, the above equation tells us that the N isotopic composition of **Acacia** is equal to the proportion of Nitrogen resulting from fixation *plus* the Nitrogen from non-fixed sources. These source terms are weighted by the isotopic composition of the atmosphere, $\delta^{15}N$, and that of the soil $\delta^{15}N$, respectively. These isotopic compositions are themselves altered by fractionation within their respective systems. The capital delta $\Delta^{15}$ terms refer to the fractionation inherent to Nitrogen fixation ("assim-atm") and nitrification ("assim-soil").

We solve for $pN_{fix}$ and simplify the above equation by assuming that $\delta^{15}N_{soil}$ + $\Delta^{15}N_{assim-soil}$ = **Aerva** $\delta^{15}N$ values. By definition, $\delta^{15}N_{atm} = 0 ‰$.

$$
pN_{fix} = \frac{\delta^{15}N_{acacia} - \delta^{15}N_{aerva}}
                {\Delta^{15}N_{assim-atm} -\delta^{15}N_{aerva}}
$$

If we assume a value for $\Delta^{15}N_{assim-atm}$, we can use this equation to estimate pNfix values for each pair of Acacia and Aerva samples. For this model, we will assume $\Delta^{15}N_{assim-atm} = 0‰$ i.e. we assume that there is no fractionation between atmospheric nitrogen and the fixed N pool. In a second scenario, we could consider a kinetic isotope effect of biological nitrogen fixation resulting in a fractionation of -2‰, $\Delta^{15}N_{assim-atm} = -2‰$, as is commonly reported in literature.

What we are left with is the following:

$$
pN_{fix} = \frac{\delta^{15}N_{acacia} - \delta^{15}N_{aerva}}
                {- \delta^{15}N_{aerva}}
$$
As an exmaple, we can use two isotopic values for **Acacia** and **Aerva** and compute that the proportion of Nitrogen sourced from fixation is about 81%.
```{r}
d15N_acacia_example = + 0.75
d15N_aerva_example = + 4
pNfix_example = (d15N_acacia_example - d15N_aerva_example)/(-d15N_aerva_example)
pNfix_example
```

If we assume that $\Delta^{15}N_{assim-atm}$ is -2‰, we observe a different value:
```{r}
d15N_acacia_example = + 0.75
d15N_aerva_example = + 4
pNfix_example = (d15N_acacia_example - d15N_aerva_example)/(-2 - d15N_aerva_example) # fractionation of -2 permil between atmosphere and fixed N
pNfix_example
```

# Visualizing Landscape Change with 3D Plotting
Let's switch to Python:

```{python}
import numpy as np
from numpy import genfromtxt
import matplotlib.pyplot as plt
print("Hello world! Libraries are loaded.")

```

We can use `genfromtxt` to read our csv as a numpy array:
```{python}
uni_landscape_end = genfromtxt('TeamTermite/dfs/degrades_uniform/uniform_200_500_quantity_end.csv', delimiter=",")
uni_landscape_start = genfromtxt('TeamTermite/dfs/degrades_uniform/uniform_200_500_quantity_start.csv', delimiter=",")
uni_landscape_start

```
We can define a function to create a 2d heatmap that shows the location of termite mounds and the relative amount of food on such mounds. The below plot is an example of the "start" of a simulation where termite mounds are rich in food, and mounds are evenly spaced.

```{python}
def heatmap2d(arr: np.ndarray):
    plt.imshow(arr, cmap='viridis')
    plt.colorbar()
    plt.show()

heatmap2d(uni_landscape_start)
```

Here we define a 3D plotting tool and observe what the landscape food quantities look like before any herbivores have entered the arena:
```{python}
def surface3d(arr: np.ndarray):
    ny, nx = arr.shape
    xd = np.linspace(0, 1, nx)
    yd = np.linspace(0, 1, ny)
    xv, yv = np.meshgrid(xd, yd)
    fig = plt.figure()
    ax = fig.add_subplot(111, projection = '3d')
    ax.plot_surface(xv, yv, arr, cmap = 'viridis')
    plt.show()
    
```

Here is the landscape food quanitity after herbivores have passed through and eaten:
```{python}
surface3d(uni_landscape_end)
```


Here we can see the distribution in plant matter change due to grazing behavior.
```{python}
combined_uni = list([np.ndarray.flatten(uni_landscape_start), np.ndarray.flatten(uni_landscape_end)])

fig1, ax1 = plt.subplots()
ax1.violinplot(combined_uni)
ax1.set_title('Plant Population Density')
ax1.set_ylabel('Plant Matter (value per square)')
ax1.set_xticks([1,2])
ax1.set_xticklabels(['Before Grazing', 'After Grazing'])
plt.show()
```


# Creating an Isoscape
The simplest scenario to consider is that the termite mounds all adopt the maximum $\delta^{15}N$ values reflecting measured **Acacia** leaf samples. Correspondingly, the space between the termite mounds all adopt the minimum **Acacia** leaf $\delta^{15}N$ values.

$\delta^{15}N$ Acacia on Mound = + 4‰
$\delta^{15}N$ Acacia off Mound = + 0.5‰.

We can now convert our landscapes to isoscapes by assigning these values based on the food concentration at the start of the simulation.

```{python}
d15_uni_start = uni_landscape_start.copy()
d15_uni_start[d15_uni_start == 100] = 4 # delta 15N in permil
d15_uni_start[d15_uni_start == 20] = 0.5 # delta 15N in permil

d15_uni_end = uni_landscape_start.copy()
d15_uni_end[d15_uni_end == 100] = 4 # delta 15N in permil
d15_uni_end[d15_uni_end == 20] = 0.5 # delta 15N in permil

```

View the beginning of the simulation as an isoscape:
```{python}
ny, nx = d15_uni_start.shape
xd = np.linspace(0, 1, nx)
yd = np.linspace(0, 1, ny)
xv, yv = np.meshgrid(xd, yd)
p_d15_uni_start = plt.figure()
p_d15_uni_start = p_d15_uni_start.add_subplot(111, projection = '3d')
p_d15_uni_start.plot_surface(xv, yv, d15_uni_start, cmap = 'viridis')
p_d15_uni_start.set_zlabel("$\delta$$^{15}$ N ‰")
p_d15_uni_start.set_title("Start of Simulation Isoscape")
plt.show()
```

View the end of the simulation as an isoscape:
```{python}
ny, nx = d15_uni_start.shape
xd = np.linspace(0, 1, nx)
yd = np.linspace(0, 1, ny)
xv, yv = np.meshgrid(xd, yd)
p_d15_uni_end = plt.figure()
p_d15_uni_end = p_d15_uni_end.add_subplot(111, projection = '3d')
p_d15_uni_end.plot_surface(xv, yv, d15_uni_end, cmap = 'viridis')
p_d15_uni_end.set_zlabel("$\delta$$^{15}$ N ‰")
p_d15_uni_end.set_title("End of Simulation Isoscape")
plt.show()
```

Wait, aren't these plots identical?
Yes -- while the **amount** of food in and outside of the termite mounds has changed, the mean isotopic composition of these areas has not. What **has** changed is the relative abundance of these isotopes in the landscape, as animals have preferentially grazed on the "heavy" Nitrogen areas of the termite mounds. To visualize this, we will weight the isotopic composition of each gridsquare by the amount of food contained within.

We can create a 3 dimensional numpy array as such:
```{python}
d15N_uni_start_average = np.sum(uni_landscape_start * d15_uni_start) / np.sum(uni_landscape_start)
d15N_uni_start_average

d15N_uni_end_average = np.sum(uni_landscape_end * d15_uni_end) / np.sum(uni_landscape_end)
d15N_uni_end_average
d15N_uni_start_average - d15N_uni_end_average
```

In this simulation we see a decrease in the overall isotopic composition of the landscape by 0.176‰


# Comparison of landscape type

In the original model simulations, agents were released into landscapes with termite mounds spaced in either a **uniformly** or **random** arrangement. How does grazing affect the isoscape?

```{python}
rand_landscape_end = genfromtxt('TeamTermite/dfs/degrades_random/random_200_500_quantity_end.csv', delimiter=",")
rand_landscape_start = genfromtxt('TeamTermite/dfs/degrades_random/random_200_500_quantity_start.csv', delimiter=",")

heatmap2d(rand_landscape_start)
```
```{python}
surface3d(rand_landscape_start)
```

Convert landscape to isoscape as done above:
```{python}
d15_rand_start = rand_landscape_start.copy()
d15_rand_start[d15_rand_start == 100] = 4 # delta 15N in permil
d15_rand_start[d15_rand_start == 20] = 0.5 # delta 15N in permil

d15_rand_end = rand_landscape_start.copy()
d15_rand_end[d15_rand_end == 100] = 4 # delta 15N in permil
d15_rand_end[d15_rand_end == 20] = 0.5 # delta 15N in permil
```

```{python}
d15N_rand_start_average = np.sum(rand_landscape_start * d15_rand_start) / np.sum(rand_landscape_start)
d15N_rand_start_average

d15N_rand_end_average = np.sum(rand_landscape_end * d15_rand_end) / np.sum(rand_landscape_end)
d15N_rand_end_average

d15N_rand_start_average - d15N_rand_end_average
```

This grazing behavior results in a change in $\delta^{15}N$ of 0.195‰


# A more complex isoscape
Acacia $\delta^{15}N$ isoscape can be generated. Surface values are derived from the best-fit model for Acacia $\delta^{15}N$ using averages of relative latitude and longitude for the mapped region:

$$
log(\delta^{15}N + 1) = 1.2608 - 0.0347(d) + 0.000255(lat) - 6.0308x10^{-7}(long)^2
$$
**where** 
      d : distance from mound center
      lat : relative latitude
      long : relative longitude

